---
title: "Module 1: Cross-Scale Interactions Module"
author: "FREC 3004 Environmental Informatics"
output: 
    html_document: default
    github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Cross-Scale Interactions Module 

This module was initially developed by Carey, C.C. and K.J. Farrell. 
13 Aug. 2017.
Macrosystems EDDIE: Cross-Scale Interactions. 
Macrosystems EDDIE Module 2, Version 1. 
module2.macrosystemseddie.org
Module development was supported by NSF EF 1702506.

**R code for students to work through the module activities A, B, and C.**

This module consists of 6 objectives. Activity A consists of Objectives 1-2,
Activity B consists of Objectives 3-4, & Activity C consists of Objectives 5-6.

## ACTIVITY A - OBJECTIVE 1

Download R packages and GLM files onto your computer.

```{r}
if (!"sp" %in% installed.packages()) install.packages("sp")
```
NOTE: depending on your computer, you may get output 
that says, "There is a binary version available. Do you want to install from 
sources that need compilation? y/n" If this pops up, type 'y' (without the 
quotes) and hit enter. You may now be prompted to download the command line 
developer tools in a pop-up window. Command line developer tools is a program 
used to run modeling software. Click Install and then re-run the 
install.packages(sp) once the install of the tools is finished. This should 
now successfully load- when it's done, it should say 'DONE(sp)' if it worked.

```{r}
if (!"devtools" %in% installed.packages()) install.packages("devtools")
``` 
This is another R package used to run modeling 
software. If you get an error message that says, "package ‘devtools’ is not 
available (for R version x.x.x)", be sure to check that your R software is up 
to date to the most recent version.

load the packages you just downloaded
```{r}
library(sp)
library(devtools)
```
Download the GLMr software. This may take a few minutes. 
If downloaded successfully, you should see "DONE (GLMr)" at the 
end of the output.

```{r}
if (!"GLMr" %in% installed.packages()) devtools::install_github("CareyLabVT/GLMr", force = TRUE)
```

This step downloads the R packages that allow you to work with GLM in R. 

```{r}
if (!"glmtools" %in% installed.packages()) devtools::install_github("CareyLabVT/glmtools", force = TRUE)
```

Load the two packages that you need to analyze GLM output

```{r}
library(glmtools)
library(GLMr)
```

NOTE: you may get lots of output messages in red at this step- if this worked 
successfully, you should read a lot of text that starts with: "This 
information is preliminary or provisional..." 

If this worked, GLMr should load without error messages. Hooray!

See what version of GLM you are running- should be v.2.x.x

```{r}
glm_version()
```

*CONGRATS!* You've now succesfully loaded GLM onto your computer! 

Now, we will explore the files that come with your downloaded GLM files 

**NOTE!**
Throughout the rest of the module, you will need to modify some of the 
lines of code to run on your computer. If you need to modify a line, I put the 
symbols ##!! before that line's annotation.  If you do not see those 
symbols, you do not need to edit that line of code and can run it as written.

When working in R, we set the sim_folder to tell R where your files, scripts, 
and model output are stored.

To find your folder path, navigate to the 'cross_scale_interactions' folder on 
your Desktop. Right click on the folder that matches your 
model lake (Mendota or Sunapee), then select Properties (Windows) or 
Get Info (Mac). Look under Location (Windows) or Where (Mac) to find your 
folder path (examples below):

Windows: C:/Users/KJF/Desktop/cross_scale_interactions/LakeName

Mac: Users -> careylab -> Desktop -> cross_scale_interactions -> LakeName

to define the sim_folder location for your model lake. You will need to change 
the part after Users/ to give the name of your computer (e.g., my computer name 
is cayelan, but yours will be different!) AND change the word LakeName to be 
the name of your model lake (Mendota or Sunapee).

```{r}
##!!Edit this
sim_folder <- '/Users/quinn/Dropbox/Teaching/Environmental Informatics/FREC3004_EnvInformatics_Class/Module1_MacroSystems_EDDIE_CrossScale/Mendota/'
```

This line of code is used to reset your working directory to the sim_folder. 
The point of this step is to make sure that any new files you create 
(e.g., figures of output) end up together in this folder.

```{r}
setwd(sim_folder)
```

This step sets the nml_file for your 
simulation to be in the new sim_folder location.

```{r}
baseline_nml_file <- paste0(sim_folder,"/glm2_baseline.nml")
nml_file <- paste0(sim_folder,"/glm2.nml")
file.copy(baseline_nml_file,nml_file, overwrite = TRUE)
```

Read in your nml file from your new directory

```{r}
nml <- read_nml(nml_file) 
```

This shows you what is in your nml file.  This is the 'master script' 
of the GLM simulation; the nml file tells the GLM model all of the initial 
 conditions about your lake, how you are defining parameters, and more - this 
is a really important file! There should be multiple sections, including 
glm_setup, morphometry, meteorology, etc.

```{r}
print(nml)
```

Plot the meterological input data for the 
simulation: short wave & long wave radiation, air temp, relative humidity, 
etc. for the duration of the simulation. 

```{r}
plot_meteo(nml_file)
```

## ACTIVITY A - OBJECTIVE 2

Now, the fun part- we get to run the model and look at output!

```{r}
run_glm(sim_folder, verbose=TRUE)
```
So simple and elegant... if this works, you should see output that 
says "Simulation begins.." and then shows all the time steps. 
At the end, it should say "Run complete" if everything worked 
ok. This may take a few minutes.

We need to know where the output data from your simulation (the output.nc file) 
is so that the glmtools package can plot and analyze the model output. We tell 
R where to find the output file using the line below:

```{r}
#This says that the output.nc file is in the sim_folder.  
baseline <- file.path(sim_folder, 'output.nc') 
```

Plot your simulated water temperatures in a heat map, 
where time is displayed on the x-axis, lake depth 
is displayed on the y-axis, and the different colors represent different 
temperatures. 

```{r}
plot_temp(file=baseline, fig_path=FALSE)
```

To copy your plot (e.g., onto a PowerPoint slide), right click on the plot.
and click "copy image". You can then paste your plot into Word, PowerPoint, etc. 

This pair of commands can be used to list the variables that were output as part 
of your GLM run.

This will print a list of variables that the model simulates.
```{r}
var_names <- sim_vars(baseline)
print(var_names) 
```

We are particularly interested in the amount of total chlorophyll-a (chl-a), 
because that is related to phytoplankton blooms. The variable name for chl-a 
is "PHY_TCHLA", and it is reported in units of micrograms per liter of water 
(ug/L). Search through the list of variables to find PHY_TCHLA.

Use the code below to create a heatmap of chl-a in the lake over time. 

```{r}
plot_var(file = baseline, "PHY_TCHLA")
```

What do you notice about seasonal patterns in chl-a? 

We also want to save the model output of the daily chlorophyll-a concentrations 
in the lake during our baseline simulation, because we'll be comparing it to 
our climate and land use scenarios later.  To do this, we use the following 
commands:

Save the chl-a from the surface only:

```{r}
chla_output <- get_var(file=baseline, "PHY_TCHLA", reference='surface', z_out=c(1)) 
```

Here we rename the chl-a column so we remember it is from the Baseline scenario:

```{r}
colnames(chla_output)[2] <- "Baseline_Chla" 
```

## ACTIVITY B - OBJECTIVE 3

For Activity B, you will work with your partner to model your lake, plus another 
team that is modeling another lake. With your partner and another team, select
one of the pre-made climate scenarios. Remember that both teams should run the 
SAME climate scenario on their separate lakes and compare the output.

Once you have selected your climate scenario, you need to edit the glm2_baseline.nml file 
to change the name of the input met file so that it reads in the met data for 
our climate scenario, not the default 'met_hourly.csv'.  

Open the .nml file by clicking 'glm2_baseline.nml' in the Files tab of RStudio, then scroll 
down to the meteorology section, and change the 'meteo_fl' entry to the new 
met file name (e.g., from 'met_hourly.csv' to 'met_hourly_plus2.csv'). 

USE **SAVE AS** to save your modified 'glm2_baseline.nml' file as 'glm2_climate.nml'

Once you have edited the nml file name, you should check to make sure that 
it is correct with the following commands:

```{r}
climate_nml_file <- paste0(sim_folder,"/glm2_climate.nml")
nml_file <- paste0(sim_folder,"/glm2.nml")
file.copy(climate_nml_file, nml_file, overwrite = TRUE)
nml <- read_nml(nml_file)
```

The printout here should list your NEW meteorological file for your climate 
scenario. If it doesn't, make sure you pressed the Save icon (the floppy disk)
after you changed your glm2.nml file.

```{r}
get_nml_value(nml, 'meteo_fl') 
```

You can now run the model for your climate change scenario using the new edited 
nml file using the commands below. Exciting!

```{r}
run_glm(sim_folder, verbose=TRUE) # Run your GLM model for your lake climate scenario. 
```

Again, we need to tell R where the output.nc file is so that the glmtools package 
can plot and analyze the model output. We tell R where to find the output file 
using the line below:

```{r}
climate <- file.path(sim_folder, 'output.nc')
```

This defines the output.nc file 
as being within the sim_folder. Note that we've called this output "climate" 
since it is the output from our climate change simulation.

As before, we want to save the model output of the daily chlorophyll-a 
concentrations in the lake during our climate change simulation, to compare to 
our baseline and land use scenarios later. 

Extract surface chl-a:

```{r}
climate_chla <- get_var(file=climate, "PHY_TCHLA", reference='surface', z_out=c(1)) 
chla_output["Climate_Chla"] <- climate_chla[2]
```

Here we attach the chl-a data from your 
climate simulation to the same file that contains your baseline scenario chl-a 
concentrations. 

To check that your climate change scenario ran correctly, run the command below, and
compare the chl-a data between your baseline and climate scenarios. They'll likely 
be similar, but if they're EXACTLY the same after the first few rows, something might 
have gone wrong in setting up your climate scenario (likely with changing the glm2.nml file!)

```{r eval = FALSE}
View(chla_output)
```

## ACTIVITY B - OBJECTIVE 4

Plot the output using the commands you learned above.
```{r}
plot_temp(file=climate, fig_path=FALSE)
```

Create a heatmap of the water temperature.

How does this compare to your baseline?

Note: If you want to control the maximum value of the color scale on your heatmaps, 
add the following (without quotes) after fig_path=FALSE: 'col_lim= c(0,35)'
This tells R that you want your maximum value to be 35, and your min. to be 0

```{r}
plot_var(file=climate, "PHY_TCHLA") # Create a heatmap of chlorophyll-a. How 
```

How does this compare to your baseline?
You can add the 'col_lim' command to this plot, too! Look at the Note from the 
plot_temp() command to learn how.

Do these plots from the climate scenario and the baseline support or contradict 
your hypotheses about climate change effects on chlorophyll-a? 

## ACTIVITY C - OBJECTIVE 5

Now, with your partner and the other team, select one of the pre-made land use 
scenarios based on changes in phosphorus concentrations in the inflow file. 
Remember that both teams should run the SAME land use scenario on their separate 
lakes and compare the output.

Once you have selected your land use scenario, you need to edit the glm2_baseline.nml 
file to change the name of the inflow file so that it reads in the inflow file 
for your land use scenario, not the default "inflow.csv".  

Open the glm2_baseline.nml file, scroll down to the inflows section, and change the 
inflow_fl entry to the new file name (e.g., from 'inflow.csv' to 'inflow_fourP.csv'). 

USE **SAVE AS** to save your modified 'glm2_baseline.nml' file as 'glm2_landuse.nml'

Once you have edited the nml file name, check to make sure that it is correct 
with the commands:

```{r}
landuse_nml_file <- paste0(sim_folder,"/glm2_landuse.nml")
nml_file <- paste0(sim_folder,"/glm2.nml")
file.copy(landuse_nml_file, nml_file, overwrite = TRUE)
nml <- read_nml(nml_file)
```

Read in your updated nml file 

```{r}
get_nml_value(nml, 'inflow_fl')
```

You should get an output that lists the name of your ALTERED inflow file.

```{r}
get_nml_value(nml, 'meteo_fl')
```
You should get an output that lists the name of your BASELINE meteorological file ('met_hourly.csv').

You can now run the model for your land use scenario using the new edited 
nml file using the commands below. Exciting!

```{r}
run_glm(sim_folder, verbose=TRUE)
```

Run your GLM model for your lake land use scenario. At the end of the model run, 
it should say "Run complete" if everything worked.

Again, we need to tell R where the output.nc file is so that the glmtools package 
can plot and analyze the model output. We tell R where to find the output file 
using the line below:

```{r}
landuse <- file.path(sim_folder, 'output.nc')
```

This defines the output.nc file as being within the sim_folder. 

Note that we've called this output "landuse" since it is the output 
from our land use change simulation.

As before, we want to save the model output of the daily chlorophyll-a 
concentrations in the lake during our land use change simulation, to compare 
to our baseline and climate scenarios later. 

Extract surface chl-a:
```{r}
landuse_chla <- get_var(file=landuse, "PHY_TCHLA", reference='surface', z_out=c(1)) 
```

```{r}
chla_output["LandUse_Chla"] <- landuse_chla[2]
```

Here we attach the chl-a data from your land use simulation to the same 
file that contains your baseline scenario and climate change 
scenario chl-a concentrations.

Plot the output of your land use scenario using the commands you learned above. 
```{r}
plot_var(file=landuse, "PHY_TCHLA")
```

Heatmap of chla. How does your phytoplankton 
heatmap look in comparison to the baseline? Be sure to check the scale of the 
color gradient representing chl-a when comparing plots!

Finally, we want to see what happens when land use and climate interact! 
Luckily, testing the combined effects of your land use and climate change 
scenarios will be pretty easy! Since we already have modified met data 
(climate scenario) and inflow data (land use scenario), we just have GLM read 
them both at once. We can do this by changing the glm2.nml file to include our 
modified files.

In the glm2_baseline.nml file, make the following TWO changes:

1) In the meteorology section, change the 'meteo_fl' entry to the met file that 
  represents your climate change scenario (e.g., 'met_hourly_plus2.csv')

2) In the inflow section, check that the 'inflow_fl' file represents your land 
  use change scenario (e.g., 'inflow_fourP.csv')

USE **SAVE AS** to save your modified 'glm2_baseline.nml' file as 'glm2_climate_landuse.nml'

Read in your updated nml file 

```{r}
climate_landuse_nml_file <- paste0(sim_folder,"/glm2_climate_landuse.nml")
nml_file <- paste0(sim_folder,"/glm2.nml")
file.copy(climate_landuse_nml_file, nml_file, overwrite = TRUE)
nml <- read_nml(nml_file)
```

If you have done this correctly, you should get an output 
that lists the name of your ALTERED inflow file.

```{r}
get_nml_value(nml, 'inflow_fl')
```

If you have done this correctly, you should get an output that 
lists the name of your ALTERED meteorological file.

```{r}
get_nml_value(nml, 'meteo_fl')
```

Run GLM one more time!

```{r}
run_glm(sim_folder, verbose=TRUE)
```

Run your GLM model for your lake climate + land use scenario 

As above, we need to tell R where the output.nc file is:

```{r}
climate_landuse <- file.path(sim_folder, 'output.nc')
```

This defines the output.nc 
file as being within the sim_folder. Note that we've called this output 
"climate_landuse" since it is the output from our simultaneous climate AND 
land use change simulations.

As before, we want to save the model output of the daily chlorophyll-a 
concentrations in the lake, to compare to our baseline, climate, and land use 
scenarios. 

Extract surface chl-a:

```{r}
combined_chla <- get_var(file=climate_landuse, "PHY_TCHLA", reference='surface', z_out=c(1)) 
chla_output["Climate_LandUse_Chla"] <- combined_chla[2]
```

Here we attach the chl-a data from your combined simulation to the 
same file that contains your baseline, climate change, 
and land use scenario chl-a concentrations

Plot the output of your land use scenario using the commands you learned above. 

```{r}
plot_var(file=climate_landuse, "PHY_TCHLA") # Heatmap of chlorophyll-a
```

Now that you've run four different scenarios (baseline, climate only, land use 
only, and climate + land use), let's plot how the chl-a in the lakes responded 
to the different scenarios. We can do this by:

The command below plots DateTime vs. Observed data in black:

**Note** that the command ylim=c(0, 100) tells R what you want the minimum and 
maximum values on the y-axis to be (here, we're plotting from 0 to 100 ug/L). 
You should adjust this range to make sure all your data are shown in the 
plot.

```{r}
attach(chla_output)
plot(DateTime, Baseline_Chla, type="l", lwd=2, col="black", ylim=c(0, 100),
     ylab="Chlorophyll-a (ug/L)", xlab="Date")  
lines(DateTime, Climate_Chla, lwd=2, col="firebrick") # add a red line of the climate change scenario
lines(DateTime, LandUse_Chla, lwd=2, col="deepskyblue") # add a blue line of the land use scenario
lines(DateTime, Climate_LandUse_Chla, lwd=2, col="darkgreen") # add a green line of the climate + land use scenario
legend("topleft",c("Baseline", "Climate Only", "Land Use Only", "Combined C + LU"),  # add a legend
       lty=1, lwd=2, col=c("black","firebrick","deepskyblue", "darkgreen"),bty='n')
```

Now we can analysis the output for the four scenarios to check for non-linear interactions

First, we need to pick a day to examine.  Lets pick September 12, 2013, which is day 225.

```{r}
focal_day <- 225
```

Now redo the plot above with the focal date highlighted as a vertical line

```{r}
attach(chla_output)
plot(DateTime, Baseline_Chla, type="l", lwd=2, col="black", ylim=c(0, 100),
     ylab="Chlorophyll-a (ug/L)", xlab="Date")  
lines(DateTime, Climate_Chla, lwd=2, col="firebrick") # add a red line of the climate change scenario
lines(DateTime, LandUse_Chla, lwd=2, col="deepskyblue") # add a blue line of the land use scenario
lines(DateTime, Climate_LandUse_Chla, lwd=2, col="darkgreen") # add a green line of the climate + land use scenario
legend("topleft",c("Baseline", "Climate Only", "Land Use Only", "Combined C + LU"),  # add a legend
       lty=1, lwd=2, col=c("black","firebrick","deepskyblue", "darkgreen"),bty='n')
abline(v = as.POSIXct("2013-09-12"))
```

Now examine the ouput from that day to answer question 15
```{r}
chla_output[focal_day,]
```

## ACTIVITY C - OBJECTIVE 6

Using the line plot you just created, and the other team's line plot from their 
lake, put together a brief presentation of your model simulation and output to 
share with the rest of the class (you'll present as a group of 4!)

Make sure your presentation answers the questions listed in your handout.

Bravo, you are done!! 

## Feedback

We welcome feedback on this module and encourage you to provide comments, 
questions, and suggestions. 

Please visit our website (http://MacrosystemsEDDIE.org) to submit feedback to 
the module developers.