---
title: "Working with strings and writing functions"
author: 'FREC 3004: Environmental Informatics'
date: "3/20/2020"
output:
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#Working with strings

Strings are set of characters that are combined into a single element of a 
vector or array. For example, the following vector has three strings

```{r}
string_vector <- c("20190831", "20190901", "20190902")
string_vector
```

Often there is information within a string that you want to analyze or use.  For
example, the object `string_vector` is a vector of dates in a format that is not
commonly used.  We are interested in separating the components of each string so
that we can figure out which year, month, and day the string is trying to 
represent.

Fortunately, there is a set of functions that we can use in R to extract 
information or modify strings in the `stringr` package that is part of the 
tidyverse.  The `stringr` cheatsheet can be found here: 

https://github.com/rstudio/cheatsheets/blob/master/strings.pdf

* We we know that the first 4 numbers are the year. We can use the 
`str_sub()` to extract the first 4 characters of the string

```{r}
str_sub(string_vector, start = 1, end = 4)
```

likewise we can extract the month using

```{r}
str_sub(string_vector, start = 5, end = 6)
```

* There are many other functions extract particular sets of characters.  For 
example the `str_extract()` function pulls out a specified string if present.

```{r}
str_extract(string_vector, "09")
```

* You can check whether a string has a particular string within it

```{r}
str_detect(string_vector, "09")
```

* If you have a vector of string that you want to separate
into parts and you know that they are separated by specific character 
(like a space - " "), you can use the `separate` function to separate the 
string. The `sep` is the separating character and the `into` is the name of the 
columns that you want to save each part.

```{r}
species <- tibble(species = c("Picea abies",
                              "Schefflera actinophylla",
                              "Betula alleghaniensis Britton"))

split_species <- species %>% 
separate(species, sep = " ", into = c("GENUS", "SPECIES", "Other"))
split_species
```

The `str_split_fixed()` can do the same thing but it does not work as well with
piping and does not name the columns

```{r}
split_species <- str_split_fixed(species$species, pattern = " ", n = 3)
split_species
```

There are many other functions that allow you to work with and manupulate 
strings in R.  I recommend looking up the `stringr` package and the cheatsheet.

## Creating functions

While R has many functions that others have created for you to use, you will
often have a particular calculation that you want to do repeatly that does
not currently exist in an R package.  This requires you to write your own 
function.

Function creation look like this

```{r}
sum_two_numbers <- function(a, b){
  c <- a + b
  return(c)
}
```

where the function named `sum_two_numbers` and it requires to inputs 
(`a` and `b`) and returns one output (`c`).

Running the chunk of code only loads the function in memory so that it is 
aviaalble for you to use

You need to call the function to use it

```{r}
summed_numbers <- sum_two_numbers(a = 3, b = 1)
summed_numbers
```