---
title: "Multi-panel plots with Patchwork"
author: "Quinn Thomas"
date: "2/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary packages
```{r}
library(tidyverse)

```

```{r echo = FALSE}
surveys <- read_csv("/Users/quinn/Downloads/portal_data_joined.csv")

surveys_complete <- surveys %>% 
  filter(!is.na(weight),
         !is.na(hindfoot_length),
         !is.na(sex))

species_counts <- surveys_complete %>% 
  count(species_id) %>% 
  filter(n >= 50)

surveys_complete <- surveys_complete %>% 
  filter(species_id %in% species_counts$species_id)

yearly_counts <- surveys_complete %>% 
  count(year, species_id)
```

This is the same code used to build the plots in "Arranging and exporting plots"

```{r}
spp_weight_boxplot <- ggplot(data = surveys_complete,
                             mapping = aes(x = species_id, y = weight)) +
  geom_boxplot() +
  labs(x = "Species", y = "Weight (g)") +
  scale_y_log10()

spp_count_plot <- ggplot(data = yearly_counts,
                           mapping = aes(x = year, y = n, color = species_id)) +
  geom_line() +
  labs(x = "Year", y = "Abundance")
```

Instead of using grid.arrange(), we will use the patchwork package.  The patachwork package does not require you to use a function

```{r}
library(patchwork)

combo_plot <- spp_weight_boxplot + spp_count_plot
combo_plot
```

Save the plot with

```{r eval = FALSE}
ggsave("/fig/combo_plot_abun_weight.png", combo_plot, width = 10, dpi = 300)
```

# Alternative patchwork combinations

If you want them stacked you use the divide symbol

```{r}
combo_plot <- spp_weight_boxplot / spp_count_plot
combo_plot
```


or if you want two stacked and one on the side

```{r}
combo_plot <- (spp_weight_boxplot / spp_weight_boxplot) | spp_count_plot
combo_plot
```








