---
title: "New Reshaping in Tidyverse"
author: "Quinn Thomas"
date: "1/14/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reshaping with pivot_longer and pivot_wider

Loading required packages and data

```{r}
library(tidyverse)
surveys <- read_csv("/Users/quinn/Downloads/portal_data_joined.csv")
```

In the spreadsheet lesson, we discussed how to structure our data leading to the four rules defining a tidy dataset:

1. Each variable has its own column
2. Each observation has its own row
3. Each value must have its own cell
4. Each type of observational unit forms a table

Here we examine the fourth rule: Each type of observational unit forms a table.

In `surveys`, the rows of `surveys` contain the values of variables associated with each record (the unit), values such as the weight or sex of each animal associated with each record. What if instead of comparing records, we wanted to compare the different mean weight of each species between plots? (Ignoring `plot_type` for simplicity).

We’d need to create a new table where each row (the unit) is comprised of values of variables associated with each plot. In practical terms this means the values of the species in `genus` would become the names of column variables and the cells would contain the values of the mean weight observed on each plot.

Having created a new table, it is therefore straightforward to explore the relationship between the weight of different species within, and between, the plots. The key point here is that we are still following a tidy data structure, but we have *reshaped* the data according to the observations of interest: average species weight per plot instead of recordings per date.

The opposite transformation would be to transform column names into values of a variable.

We can do both these of transformations with two `tidyr` functions, `pivot_longer()` and `pivot_wider()`.

## Pivot data long to wide

`pivot_wider()` takes three principal arguments:

1. the data
2. the `names_from` column variable whose values will become new column names.
3. the `values_from` column variable whose values will fill the new column variables.

Further arguments include fill which, if set, fills in missing values with the value provided.

Let’s use `pivot_wider()` to transform surveys to find the mean weight of each species in each plot over the entire survey period. We use `filter()`, `group_by()` and `summarise()` to filter our observations and variables of interest, and create a new variable for the `mean_weight`. We use the pipe as before too.

```{r}
surveys_gw <- surveys %>%
  filter(!is.na(weight)) %>%
  group_by(genus, plot_id) %>%
  summarize(mean_weight = mean(weight))
str(surveys_gw)
```

This yields `surveys_gw` where the observations for each plot are spread across multiple rows, 196 observations of 3 variables. Using `pivot_wider()` with `names_from` from genus with `values_from` from `mean_weight` this becomes 24 observations of 11 variables, one row for each plot. We again use pipes:

```{r}
surveys_wider <- surveys_gw %>%
  pivot_wider(names_from = genus, 
              values_from = mean_weight)
str(surveys_wider)
```

Importantly, when using `pivot_wider()` the `names_from` value will be an existing column name in the data frame.  Therefore it does not have quotes around it (i.e., use genus rather than "genus").  This also applies to the `value_from` arguement.

We could now plot comparisons between the weight of species in different plots, although we may wish to fill in the missing values first using the `values_fill` option.

```{r}
surveys_gw %>%
  pivot_wider(names_from = genus, 
              values_from = mean_weight, 
              values_fill = list(mean_weight = 0)) %>%
  head()
```

## Pivoting wide to long

The opposing situation could occur if we had been provided with data in the form of `surveys_wider`, where the genus names are column names, but we wish to treat them as values of a genus variable instead.

In this situation we are gathering the column names and turning them into a pair of new variables. One variable represents the column names as values, and the other variable contains the values previously associated with the column names.

`pivot_longer()` takes four principal arguments:

1. the data
2. the names of the columns (`cols`) we use to fill the `names_to` variable (or to drop).
3. The name of the column (`names_to`) we want to create that represents the the names of the columns that are used in `cols`
4. the values column variable (`values_to`) we wish to create and fill with values associated with the selected `cols`

To recreate `surveys_gw` from surveys_spread we would create a `names_to` called genus and `values_to` called `mean_weight` and use all columns except `plot_id` for the key variable. Here we drop `plot_id` column with a minus sign.

```{r}
surveys_longer <- surveys_wider %>%
  pivot_longer(cols = -plot_id, 
               names_to = "genus", 
               values_to = "mean_weight")

str(surveys_longer)
```

Importantly, when using `pivot_longer` the `names_to` and `values_to` arguments have quotes (i.e., "genus" rather than genus) because these columns do not exist yet. The columns in the `cols` argument are existing columns in your data frame.


Note that now the `NA` genera are included in the longer format. Widening and then lengthing can be a useful way to balance out a dataset so every replicate has the same composition.

We could also have used a specification for what columns to include. This can be useful if you have a large number of identifying columns, and it’s easier to specify what to lengthen than what to leave alone. And if the columns are in a row, we don’t even need to list them all out - just use the `:` operator!

```{r}
surveys_longer <- surveys_wider %>%
  pivot_longer(cols = Baiomys:Spermophilus, 
               names_to = "genus", 
               values_to = "mean_weight") %>% 
  head()
```

## Challenge

1. Spread the `surveys` data frame with `year` as columns, `plot_id` as rows, and the number of genera per plot as the values. You will need to summarize before reshaping, and use the function `n_distinct()` to get the number of unique genera within a particular chunk of data. It’s a powerful function! See `?n_distinct` for more.

```{r}
rich_time <- surveys %>% 
  group_by(plot_id, year) %>% 
  summarize(n_genera = n_distinct(genus)) %>% 
  pivot_wider(names_from = year, 
              values_from = n_genera)

head(rich_time)
```

2. Now take that data frame and `pivot_longer()`, so each row is a unique `plot_id` by `year` combination.

```{r}
rich_time %>% 
  pivot_longer(cols = -plot_id, 
               names_to = "year", 
               values_to = "n_genera")

```

3. The `surveys` data set has two measurement columns: `hindfoot_length` and `weight`. This makes it difficult to do things like look at the relationship between mean values of each measurement per year in different plot types. Let’s walk through a common solution for this type of problem. First, use `pivot_longer()` to create a dataset where we have a column called `measurement` and a column called `value` that takes on the value of either `hindfoot_length` or `weight`. Hint: You’ll need to specify which columns are being lengthened.

```{r}
survey_long <- surveys %>% 
  pivot_longer(cols = c("hindfoot_length","weight"), 
               names_to = "measurement", 
               values_to = "value")

```
4. With this new data set, calculate the average of each measurement in each year for each different `plot_type`. Then `pivot_wider()` them into a data set with a column for `hindfoot_length` and `weight`. Hint: You only need to specify the `names_from` and `values_from` columns for `pivot_wider()`.

```{r}
survey_long %>% 
  group_by(year, measurement, plot_type) %>% 
  summarize(mean_value = mean(value, na.rm = TRUE)) %>% 
  pivot_wider(names_from = "measurement", values_from = "mean_value")
```

