---
title: "Module 9 Carbon Stocks"
author: 'FREC 3004: Environmental Informatics'
output: 
    html_document: default
    github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(neonUtilities)
```

## Science question

How much carbon is stored in forest ecosystems and where in the ecosystem is 
it stored?

## Environmental Learning objectives

* Describe how carbon stocks vary between pools (i.e., vegetation vs. soil) and 
  ecosystem type (i.e., arctic vs. temperate forests) 
* Calculate carbon in trees from individual tree measurements of size
* Calculate carbon in soil from horizon measurements
* Scale carbon estimates from the individual tree to the plot (i.e., a set area)

## Data Science Learning objectives

* Developing a data science analysis plan
* Understands unfamiliar data tables and how they are linked.
* Applying prior data science skills to execute an analysis plan
* Generating an analysis that uses multiple tables of data
* Downloading and processing data from the National Ecological Observatory 
  Network

## Assumes knowledge of the following in R:

* mutate()
* group_by()
* ggplot()
* geom_line()
* geom_point()
* select()
* summarize()
* mean()
* filter()
* rename()
* read_csv()
* as.numeric()
* lm()
* rbind()
* geom_abline()
* geom_smooth()
* geom_histogram()
* geom_bar()
* head()
* read_delim()
* rename()
* slice()
* as.numeric()
* paste0()
* for loop
* if-else statement
* ifelse()
* vector()
* geom_hline()
* joins
* str_sub()
* add_predictions()
* add_residuals()
* make_datetime()
* creating simple functions

## R knowledge covered in module

- NEON API (zipsByProduct() and stackByTable())

## Step 1: Background on carbon and climate

Estimating carbon storage is important for quantifying the climate mitigation 
benefit provided by an ecosystem.  Carbon that is stored in an ecosystem is 
carbon that is not in the atmosphere and contributing to climate change.  Carbon
storage is quantified by adding up the carbon in different "stocks" of carbon 
within the ecosystem, with vegetation and soils being the most important stocks 
in most terrestrial ecosystems.  

Importantly, this is a different perspective than Module 6 where you analyzed 
the flux of carbon into or out of an ecosystem (NEP).  The carbon flux is the 
*rate* that carbon is stored while the carbon stocks are the amount of carbon 
that is already present. As an example, a young forest may have low carbon 
stocks but high NEP while an older forest may have high carbon stocks but low 
NEP.  

## Step 2: Background on estimating carbon stocks

**Calculating total biomass of trees**

One common approach to estimating carbon in woody vegetation is by measuring the
diameter of each individual tree over a certain size in a set area (called a 
plot).  The diameter of each tree is converted to biomass using allometric 
relationships between the diameter and biomass.  Allometric relationships are 
statistical relationships that are created by measuring a tree's diameter before
cutting down and weighing it.  A general relationship for a species can be 
created by combining these harvested trees into a single analysis.  Allometric 
relationships typically estimate aboveground biomass which needs to be converted
to carbon by multiplying by 0.5 (biomass is about 50% carbon).  

Total biomass of a tree is the sum of aboveground and belowground carbon.  Since
belowground carbon is less commonly measured (it requires digging up the tree 
roots), we can use ratios of aboveground to belowground carbon from a limited 
set of trees where the entire tree was extracted to calculate total tree tree 
carbon.  

Once total tree carbon is calculated for each tree (i.e., units of kg C per 
tree) in a plot, the density of carbon for each plot (i.e., kg C per m2) is 
calculated by summing across the individual trees and dividing the plot sum by 
the plot area.

**Calculating soil carbon**

One common method for estimating carbon in soils is by digging a soil pit.  A 
soil pit is a defined area (i.e., 1 m2) and researchers dig down to set depth 
(i.e., 1.5 m). As they dig, they divide the pit into soil horizons, measure the 
depth of each horizon, and separate the soil from each horizon.  Once separated,
they weight the mass of each horizon and calculate the bulk density, which is 
the mass of soil (i.e., kg) per volume of the horizon in final units of kg/m3.  

Finally, they save a sample of soil from each horizon and send it to a lab to 
measure the carbon in the sample.  The carbon in the sample is measured as mass 
of carbon (i.e, g C) per mass of soil (i.e., g soil). Ultimately to calculate 
the carbon in a soil pit, you need to use depth of each horizon to get the 
volume of the horizon.  Then you need to use  the bulk density of each horizon 
to estimate the mass of soil in each horizon (i.e., kg per horizon).  Then you 
multiple the mass of soil times the carbon per mass of soil (i.e., kgC/kg soil) 
to get the soil horizon carbon (i.e., kg C per horizon). By summing up the 
carbon per horizon, the carbon for the soil pit can be estimated (i.e., kg C/m2).

## Step 3: Your charge!

I am a client interested in the carbon storage of different ecosystems across 
the continental U.S. to guide my investment in the California Carbon Market.  

I need carbon stocks using measurements for the following sites within the 
National Ecological Observatory Network.  

Group 1:

* A. Blandy Experimental Farm (BLAN)
* B. Bartlett Experimental Forest (BART)
* C. Mountain Lake Biological Station (MLBS)
* D. Ordway-Swisher Biological Station (OSBS)

Group 2:

* A. Wind River Experimental Forest (WREF)
* B. Yellowstone Northern Range (Frog Rock) (YELL)
* C. Caribou-Poker Creeks Research Watershed (BONA)
* D. Abby Road (ABBY)

I will assign half the class to one group and the other half to the other.  For 
your group of four sites, your charge is to:

1. Write down and add your hypothesized answer to the following questions to the
class Google sheet. Explore the descriptions of the NEON sites at 
neonscience.org, how the sites look on Google Maps, and your general knowledge 
about ecosystem science to develop your hypotheses. You will write your rank for
each question.

   https://docs.google.com/spreadsheets/d/183I3juYv20EeYd-7xJ5ycYpJGXjxyLYRkrtWcDAw8Fg/edit?usp=sharing

   * What do you think are the patterns in total carbon stocks 
   (total vegetation + soil) among the sites that I provide you? Based on 
   information about location and ecosystem type, record what you think the 
   order of the sites is - with a rank of 1 being the most carbon.
   
   * At each site, hypothesize whether you think vegetation (both living and 
   dead) or soil carbon will be larger.
   
2. Write the code to calculate the total carbon stocks (live trees, standing 
dead trees, and soils)

3. Create a knitted R markdown document that reports the amount of carbon in 
each stock (live trees, standing dead trees, and soils). The document should 
include the following:

   * A figure showing the live tree stocks in each tower plot in each year with 
     measurements
   * A figure showing the dead tree stocks in each tower plot in each year with 
     measurements
   * A bar chart showing how the carbon is distributed among the three carbon 
     components
   * A table with summarized values for the site level carbon stocks in the 
     three carbon components.


4. Add your estimates for total vegetation and soil carbon to the Google sheet 
class document. See Step 11 below for the link.

**Important note:**  All values of carbon stocks should be in kg C per m2.  Your
final vegetation stock numbers for the site should be an average across all 
'tower' plots.

## Step 4: Examine data products

NEON data is organized by data product ID in the NEON Date Portal: <https://data.neonscience.org/static/browse.html>
 
You will be using the following data products:

* DP1.10098.001 (Woody plant vegetation structure)
* DP1.00096.001 (Soil physical properties (Megapit))
* DP1.00097.001 (Soil chemical properties (Megapit))
      
For each data product summarize the following based on the documents available 
in "View Product Details" link associated with each data product. Think about 
the answers to the following questions.

* How is the plot sampling done for each data product?
* What variables do you need from each data product to calculate the 
     carbon stocks in vegetation and soil at the site?
* What measurements are done that will allow for your to calculate 
     vegetation and soil carbon stocks - refer to Step 2 and the presentation.
* How is the data in each product organized?

## Step 5: Plan analysis

Using the description about how to calculate the vegetation and carbon stocks 
above, create a bulleted list your plan for calculating tree and soil carbon 
stocks at your site.

## Step 6: Download data

First, define the site ID

```{r}
site <- "ORNL"
```

and the data product ID 

```{r}
data_product <- "DP1.10098.001"
```

Second, use the `zipsByProduct()` function to download files via NEON's API

```{r}
zipsByProduct(dpID=data_product, site=site, 
                package="basic", check.size=F,
                savepath = NA)
```

Third, since NEON data is organized by month and each monthly data package as 
multiple data tables, NEON has created a function to combined tables across 
months.  Use this function `stackByTable()` to create tables with the full time 
series

```{r}
stackByTable(filepath= "filesToStack10098/",
             folder=T)
```

Look in the `filesToStack10098` folder for the files.  There may be multiple csv
files, xml, and README files.  You will need to open the csvs and consult the 
NEON documentation to understand what variables are in what tables.

## Step 7: Calculate carbon in live trees

This step will challenge you to develop a workflow using the data science skills
to calculate the carbon stocks in live trees at your site.  In the end you 
should have a *site-level* mean carbon stock in *live trees* for each year that 
was sampled.  Your estimate will be from the plots that are sampling the 
ecosystem under the flux tower - called "tower plots".  See https://www.neonscience.org/field-sites/field-sites-map/HARV for an example map 
of a plot with the tower plots labeled.

Hints for calculating carbon in live trees:

* Remember that individual trees are within plots that have a set area.  
  Individual trees and plots are remeasured through time.
* Use `filter(str_detect(plantStatus,"Live")` to select  for the live trees.  
  The `str_detect()` function will find the rows with "Live" anywhere in variable 
  character string.  This is necessary because there are multiple types of live 
  trees.
* Remember that only 50% of biomass is carbon so you will need to convert from 
  biomass to carbon
* Assume that belowground biomass (i.e., roots) is 30% of aboveground biomass
* Be very careful with the units at each step: the final units should be 
  kgC m^-2 (Kilogram of carbon per meter squared).  The order of magnitude 
  should be 1 - 100.  
* The information that you need is found across multiple tables.  Joining tables
  can be tricky because the plot table has repeating plotIDs, resulting in 
  duplicated values when you join.  The function `distinct(plotID, .keep_all = TRUE)` 
  can be used to select each plotID only once.
* The Climate Action Reserve project (an official carbon accounting organization
  for the California Carbon Exchange) provides allometric relationships to use.
  The allometric equations for each species can be found in the file 
  "Allometrics.csv" on modules.  Places this file in you data sub-directory. You
  will need to join the parameters in "Allometrics.csv"" to your data frame with
  the diameter measurements. If a species is not in the list of allometric 
  relationships it should have an NA when you join, just pick one set of 
  allometric parameters to to use and assign it to all species with NAs (these 
  are very uncommon species to it won't influence your answer much).  The 
  parameters we use come from Table 4 in the Jenkins et al. 2003 paper on Canvas.
* The site level value is the average across plots. Don't forget that we are 
  only interested in the "Tower" plots.
* Remember that each plot represents an area of forest where all the trees are 
  measured.  

Finally, remember to review the description in Step 2 about how to calculate 
carbon stocks.

## Step 8: Calculate carbon in dead trees

You will calculate carbon in standing dead trees using the same approach as used
for live trees, except changing what you use in the str_detect() function.

## Step 9: Calculate carbon in soils

This step will challenge you to develop a workflow using the data science skills
to calculate the carbon stocks in the soil at your site.  In the end you should 
have an estimate of soil carbon from the soil Megapit. 

In the NEON data, the bulk density is in the physical soil data product while 
measurements of the carbon in the sample are the soil biogeochemistry data 
product.  You will need to download both and join the tables.

Hints for calculating carbon in soil:

* Be sure to use the create depth variables, some cover all depths while others 
  do not
* Use the organic carbon variable rather than the total carbon
* Be careful about units and conversions
* Be careful about the type of horizon measurements.  Some are the 
  original measurement and some are remeasurements to audit the original 
  measurement. Using both will mess up your analysis.

```{r}
#Download biogeochemistry soil data to get carbon concentration
soil_biogeochem <- "DP1.00097.001"
#Download physical soil data to get bulk density
soil_physics <- "DP1.00096.001"
```

## Step 10: Create report

Again, I am looking for:

* A figure showing the live tree stocks in each tower plot in each year with 
  measurements
* A figure showing the dead tree stocks in each tower plot in each year with 
  measurements
* A bar chart showing the carbon in the three carbon components
* Summarized table of values for the site level carbon stocks in the three 
  carbon components.

Be sure your figures are complete with units, labels, and a title.  Be sure your
figures are legible when knitted.

## Step 11:  Add your calculations to the class Google sheet

Add your site-level data to the google sheet table

The link can be found here: https://docs.google.com/spreadsheets/d/1Byxiy5uLMplqDTmBIhVEYx0JScDUAzFa832lwYR4qr4/edit?usp=sharing