---
title: "Figure Expectations"
author: "Quinn Thomas"
date: "1/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load data for examples
```{r echo = FALSE}
library(tidyverse)
surveys <- read_csv("/Users/quinn/Downloads/portal_data_joined.csv")

surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex

species_counts <- surveys_complete %>%
  count(species_id) %>% 
  filter(n >= 50)

## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)
```

## Expectations for figures


- All plots must have a title
- All plots must have a x- and y-axis labels that include units.  The name must not have an underscore and must be actual words.
- If a legend is used, the title of the legend must not have an underscore and must be actual words.  

Here is an example of a figure that will lose points

```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length, color = species_id)) +
    geom_point()
```

Here is an example of a plot that would receive full credit

```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length, color = species_id)) +
    geom_point() +
  labs(x = "weight (g)", 
       y = "hindfoot length (cm)", 
       title = "Weight vs. hindfoot length by species", 
       color = "Species ID")
```

If you need to change the title of a legend, look at what symbols you are using to generate the title.  If you are using color in the `aes()` (e.g., `aes(x = weight, y = hindfoot_length, color = species_id)` then use `color` in the `labs()` function to change the legend title .  Other plots might use `fill`, `line_type`, or `group` in the `aes()` to create the legend.