---
title: "Joins"
author: 'FREC 3004: Environmental Informatics'
date: "3/20/2020"
output:   
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

First create two data frames: `ecosystem_type_table` and `bird_count`.  
Notice how they share two of the same column names (`siteID`) but 
the values in the two columns are not exactly similar.  In this example, 
`ecosystem_type_table` is a table of ecosystem types from one dataset and 
`bird_count` is the number of bird species in a set of plots.  Not all sites 
with bird species counts were in the set of sites where the ecosystem type was 
evaluated.

```{r}
ecosystem_type_table <- tibble(siteID = c("MLBS","BART","TOOK"),
            ecosystem_type = c("forest","grassland","grassland"))
ecosystem_type_table
```

```{r}
bird_count_table <- tibble(siteID = c("MLBS","MLBS", "BART","TALL"),
             plotID = c("A1", "A2", "A1", "B30"),
            bird_count = c(30,10,20,5))
bird_count_table
```

The goal of joins is to combine these two table.  However, there are different 
ways to combined tables based on the desired outcome.

## Left join

Now, you want to join the two columns so that you merge the `bird_count_table` 
table into the `ecosystem_type_table` table. The key is that you want to keep 
all rows from `ecosystem_type_table` but merge in rows from `bird_count_table` 
that match rows from `ecosystem_type_table` (i.e., the same `siteID`). To do 
this we use an left join.  The reason "left" is used is becausewe want to keep 
all rows in the left table of the function call

```{r}
left_join(ecosystem_type_table, bird_count_table, by = c("siteID"))

```

See how `siteID =  TOOK` has an `NA` for `bird_count`.  This is because `TOOK` 
is not in the `bird_count_table` table and the `bird_count_table` table was on 
the *right* side of the function call. Similarly, the `siteID = TALL` is missing
because it is only in the `bird_count_table` table. Notice how all the rows in 
the `ecosystem_type_table`.

In the example, you specified `by = c("siteID")`. If you set 
`by = NULL` then it would have automatically found the columns in the
two tables that were the same

```{r}
left_join(ecosystem_type_table, bird_count_table, by = NULL)
```

which is the same as

```{r}
left_join(ecosystem_type_table, bird_count_table)
```

## Right join

The right join is the exact opposite of the left join.  Where you want to keep
all rows from the table that is on the right side of the function call

```{r}
right_join(ecosystem_type_table, bird_count_table, by = c("siteID"))
```

Notice how this has the `siteID = TALL` which is only in the `bird_count_table` 
table. If you switch the order of the table you can get the same answer with a 
left join.

```{r}
left_join(bird_count_table, ecosystem_type_table, by = c("siteID"))
```

## Inner join

If you want to only include rows that are in both tables, you would use an 
inner join.  For example, we would use an inner join if we only want the 
`SiteID` that are in both the `bird_count_table` and 
`ecosystem_type_table` tables.  You will see inner joins referred to as 
"filtering" joins because the resulting table is a subset of the original tables.

```{r}
inner_join(ecosystem_type_table, bird_count_table, by = c("siteID"))
```

## Full join

If you combine all rows from both tables, you would use a full join.  This join
will result in the most `NA` values

```{r}
full_join(ecosystem_type_table, bird_count_table, by = c("siteID"))
```

## Issues with Joins

There can be issues when joining if you have two tables and they have columns 
with the same names but not all columns with the same names are in the `by =`. 
The result of the join will add a `.x` and `.y` to the names of the shared 
columns that are not in the `by = ` argument.  Sometimes this can incorrectly
filter or replicate your data.  If you see `.x` and `.y` in your column names,
be sure to comfirm that your join worked correctly.