---
title: "Joins"
author: 'FREC 3004: Environmental Informatics'
date: "3/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

First create two data frames: `ecosystem_type_table` and `bird_count`.  
Notice how they share two of the same column names (`siteID` and `plotID`) but 
the values in the two columns are not exactly similar.  In this example, 
`ecosystem_type_table` is a table of field plots from one dataset and 
`bird_count` is the number of bird species in a set of plots.  Not all plots 
with bird species counts were in the set of plots where the ecosystem type was 
evaluated.

```{r}
ecosystem_type_table <- tibble(siteID = c("MLBS","MLBS","BART","TOOK"),
            plotID = c("A1","A2", "A1", "B10"),
            ecosystem_type = c("forest","forest","grassland","grassland"))
ecosystem_type_table
```

```{r}
bird_count_table <- tibble(siteID = c("MLBS","BART","TALL"),
             plotID = c("A1", "A1", "B30"),
            bird_count = c(30,20,5))
bird_count_table
```

The goal of joins is to combine these two table.  However, there are different 
ways to combined tables based on the desired outcome.

## Left join

Now, you want to join the two columns so that you merge the `bird_count_table` 
table into the `ecosystem_type_table` table. The key is that you want to keep 
all rows from `ecosystem_type_table` but merge in rows from `bird_count_table` 
that match rows from `ecosystem_type_table` (i.e., the same `siteID` and 
`plotID`). To do this we use an left join.  The reason "left" is used is because
we want to keep all rows in the left table of the function call

```{r}
left_join(ecosystem_type_table, bird_count_table, by = c("siteID","plotID"))

```

See how `siteID =  TOOK` has an `NA` for `bird_count`.  This is because `TOOK` 
is not in the `bird_count_table` table and the `bird_count_table` table was on 
the *right* side of the function call. Similarly, the `siteID = TALL` is missing
because it is only in the `bird_count_table` table. Notice how all the rows in 
the `ecosystem_type_table`.

In the example, you specified `by = c("siteID","plotID")`. If you set 
`by = NULL` then it would have automatically found the columns in the
two tables that were the same

```{r}
left_join(ecosystem_type_table, bird_count_table, by = NULL)
```

which is the same as

```{r}
left_join(ecosystem_type_table, bird_count_table)
```

## Right join

The right join is the exact opposite of the left join.  Where you want to keep
all rows from the table that is on the right side of the function call

```{r}
right_join(ecosystem_type_table, bird_count_table, by = c("siteID","plotID"))
```

Notice how this has the `siteID = TALL` which is only in the `bird_count_table` 
table. If you switch the order of the table you can get the same answer with a 
left join.

```{r}
left_join(bird_count_table, ecosystem_type_table, by = c("siteID","plotID"))
```

## Inner join

If you want to only include rows that are in both tables, you would use an 
inner join.  For example, we would use an inner join if we only want the 
`SiteID` and `plotID` that are in both the `bird_count_table` and 
`ecosystem_type_table` tables.  You will see inner joins referred to as 
"filtering" joins because the resulting table is a subset of the original tables.

```{r}
inner_join(ecosystem_type_table, bird_count_table, by = c("siteID","plotID"))
```

## Full join

If you combine all rows from both tables, you would use a full join.  This join
will result in the most `NA` values

```{r}
full_join(ecosystem_type_table, bird_count_table, by = c("siteID","plotID"))
```

## Issues with Joins

There can be challenges with joins.  The most important is when you join two
tables and they have columns with the same names but not all columns with the
same names are in the `by =`.  For example, we could just join our tables by 
`siteID`.  

```{r}
left_join(ecosystem_type_table, bird_count_table, by = c("siteID"))
```

This results is two columns for `plotID`: `plotID.x` is from the left table and 
`plotID.y` is from the right table.  Notice how there are two rows with the same
`bird_count` value (`30`).  This result incorrectly repeated measurements 
because two different `plotIDs` share the same `siteID` and we only joined by 
`siteID`.  Be careful about joins and double check that you results are what you
expect - do not assume it worked correctly.
