---
title: 'Module 6: Terrestrial Carbon Flux'
author: "FREC 3004 Environmental Informatics"
output:  
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


```{r message= FALSE}
library(tidyverse)
library(lubridate)
library(modelr)
```
## Science question

How much carbon dioxide does a forest remove from the atmosphere?

## Environmental Learning objectives

Having completed the module students will be able to:

* Import and analyze net carbon fluxes from the FLUX2015 database using R, 
  tidyverse, and ggplot.
* Describe how carbon flux varies from minutes to annual time scales.
* Compare how the rate of carbon uptake by ecosystems varies among locations.
* Separate a net flux (NEP) into components (GPP and RE).
* Create informative figures in ggplot that display temporal variation in carbon
  fluxes.

## Data Science Learning objectives

* Create simple functions in R
* Modify and create strings

## Assumed R knowledge

* mutate()
* group_by()
* ggplot()
* geom_line()
* geom_point()
* select()
* summarize()
* mean()
* filter()
* rename()
* read_csv()
* as.numeric()
* lm()
* rbind()
* geom_abline()
* geom_smooth()
* geom_histogram()
* head()
* read_delim()
* rename()
* slice()
* as.numeric()
* paste0()
* for loop
* if-else statement
* ifelse()
* vector()
* geom_hline()
* joins

## R knowledge covered in module

* str_sub()
* add_predictions()
* add_residuals()
* make_datetime()
* creating simple functions

## Background, and why you should care: 

Carbon is the currency of ecosystem production, comprising half of the dry 
weight of plant tissues produced when atmospheric CO2 is fixed via 
photosynthesis and then used to build biomass. The rate at which new plant 
biomass accumulates or, put another way, the “flux” of carbon from the 
atmosphere to plant biomass is net primary production (NPP). A closely related 
term, net ecosystem production (NEP), is the balance between gross primary 
production (GPP, whole-canopy carbon fixation) and ecosystem respiration (Re, 
respiration from all organisms). Often, NEP is equated with “carbon 
sequestration”, a general term used to describe the quantity of carbon dioxide 
removed from the atmosphere, because it is a measure of the net balance between 
CO2 moving into the forest from photosynthetic carbon fixation and that which is
released back to the atmosphere from respiration. This balance between whole-
ecosystem carbon uptake (i.e., GPP) and loss (i.e., Re) changes over time as 
leaves grow and later fall, as shifting climate alters the balance between these
two opposing fluxes of carbon, and as disturbances modify the number and 
physiology of primary producers. Spatially, variability in climate and 
disturbance regime may explain why ecosystems in different regions sequester 
carbon at different rates. Understanding how much carbon ecosystems sequester 
and why is therefore a question of fundamental importance to ecologists, and 
also to the management of terrestrial ecosystems for carbon sequestration. 

Facilitating the quest to answer these questions, researchers are increasingly 
drawing from open research networks that provide high temporal resolution 
(usually half-hourly to hourly) estimates of NEP, GPP, and Re for different 
ecosystems all over the world. Among these networks, the National Ecological 
Observatory Network (NEON, https://www.neonscience.org/) and FLUXNET 
(http://fluxnet.fluxdata.org/) provide freely available carbon flux data. When 
combined with long-term environmental data from networks collecting 
meteorological information such as the National Oceanic and Atmospheric 
Administration (NOAA, http://www.noaa.gov/), it is possible to assess the 
sensitivity of carbon fluxes to ongoing and predicted climate change. This lab 
challenges participants to use open, long-term data from a variety of networks 
to ask: 

1) How variable are carbon fluxes among sites and over time?; 
2) Do temporal patterns of carbon flux differ at different timescales of minutes, 
days, and years?; and 
3) Given historical climate trends, how will changes over 
time, should they continue, affect the future trajectory of carbon fluxes?

## Part 1: Site selection and hypothesis development

Step 1: Select a unique NEON eco-region to investigate (Figure 1). You can read 
about the different NEON eco-regions and how they were selected here: 
https://www.neonscience.org/field-sites.

Step 2: Within your selected NEON eco-region, choose a site to investigate
  
  1) Go to <http://fluxnet.fluxdata.org>
  2) Go to the "Site" tab, click on "Site List", and choose "Map View".
  3) Click on sites in your eco-region and click on different sites to find one 
    that has "Data Start" and "Data End" dates that span 5 or more years
  4) Take note of the name of the site (i.e., US-Ha1)
  
**Question 1:** What site did you pick?  What ecoregion is it in?

**Answer 1:**

Step 3: Develop hypotheses for your site. Specifically, describe what you think 
temporal patterns of NEP, GPP, and RE at the following temporal scales will be: 

**Question 2:** 

Hourly within a day

**Answer 2:**

* NEP:
* GPP:
* RE:

**Question 3:** 

Monthly to month

**Answer 3:**

* NEP:
* GPP:
* RE:

**Question 4:** 

Year to year

**Answer 4:**

* NEP:
* GPP:
* RE:

## Part 2: Test your hypotheses using real data

Step 1: Download carbon flux data for your site.

1) Go to <http://fluxnet.fluxdata.org>
2) Click "Sign-In" and create an account using your Virginia Tech email
3) Sign in to the site
4) Click "Data" tab and chose the "FLUXNET2015 Dataset"
5) Click "Download FLUXNET2015 Dataset"
6) Chose "Tier 1 Data"
7) Click "FULLSET data product (includes SUBSET)"
8) In the list of Site IDs, chose your Site and move to the "Your Selections"
9) Chose "Education" for the "Intended Use"
10) In the description, write "For analysis in Virginia Tech FREC 3004 
   Environmental Informatics course"
11) Open the data policy and summarize how it applies to use in this course

**Question 5:** How are we meeting the data policy requirements?

**Answer 5:**

12) If we are meeting the data policy requirements, click "I acknowledge.."
13) Click "Download All Files"
14) Click on the link with your Site ID
15) Move the files to your module data subfolder

Step 2: Import the hourly time-step data (i.e., file with 
“FLUXNET2015_FULLSET_HH”). Orient and familiarize yourself with the variable 
headers: 
http://fluxnet.fluxdata.org/data/fluxnet2015-dataset/fullset-data-product/. 

```{r}
#INSERT CODE HERE
```

Step 3: Tidy up the timedate variable.  The timestep variable is a non-standard 
format that we need to clean up using a new set of functions.  First, notice the
format of `TIMESTAMP_START`.  It will vary depending on your site but in general
it will look like this: 199101010000.  The first 4 numbers are the year, next 2 
are the month, next 2 are the day, next 2 are the hour, and next 2 are the 
minute.  We can use this information along with the function `str_sub()` in the
`stringr` package that comes with `tidyverse` to grab the correct set of numbers
for the corresponding unit of time.  Here is an example:  The numbers in the 
str_sub() below are the beginning and ending character to grab.

```{r}
str_sub("EI_IS_GREAT", 1, 2)
str_sub("EI_IS_GREAT", 7, 11)
```

Since it is a character it needs to be converted to a number using `as.numeric`.
Then you can use the `make_datetime()` function that is in the `lubridate`
package to combine. 
 
```{r}
flux_data <- flux_data %>% 
  mutate(YEAR = as.numeric(str_sub(TIMESTAMP_START, 1, 4)),
         MONTH = as.numeric(str_sub(TIMESTAMP_START, 5, 6)),
         DAY = as.numeric(str_sub(TIMESTAMP_START, 7, 8)),
         HR = as.numeric(str_sub(TIMESTAMP_START, 9, 10)),
         MIN = as.numeric(str_sub(TIMESTAMP_START, 11, 12)),
         TIMESTAMP_START = make_datetime(YEAR, MONTH, DAY, HR, MIN, tz = "UTC")) %>% 
  select(TIMESTAMP_START, YEAR, MONTH, DAY, HR, MIN, NIGHT, NEE_VUT_REF, TA_F)
```

The flux towers measure Net Ecosystem Exchange (NEE), which is negative when the
system takes up carbon from the atmosphere and positive when it releases carbon.
First, check to see what value is used to represent NA values in the NEE_VUT_REF
variable.  Filter out NA values. Be sure that your site still has 5-years of 
data after filtering out NA values.  If not, pick a new site and acquire that 
data. Second, we are going to analyze Net Ecosystem Productivity (NEP), which is
the negative of NEE. NEE is NEE_VUT_REF in the data set.  Create a new column 
called NEP that is -NEE_VUT_REF. 

A positive value for NEP means carbon is being absorbed from the atmosphere 
because GPP > RE and a negative value means that NEP is being released by the 
ecosystem to the atmosphere because RE > GPP. 

Note that the units of NEP are umolCO2 /m2 /second.

```{r}
#INSERT CODE
```

Step 4: Given your hypotheses, consider how to organize and then plot your data. 
For examining how carbon fluxes change from month-to-month, what is your x-axis?
What about your y-axis? How do you organize your data?

Plot the entire hourly time-series for NEP and describe the patterns.  Be sure 
to add a horizontal line at 0 so that it is easy to see when NEP is positive or
negative.

```{r}
#INSERT CODE
```

Plot the time-series for NEP at annual scale (calculate the annual mean).  Be 
sure to add a horizontal line at 0 so that it is easy to see when NEP is 
positive or negative.

```{r}
#INSERT CODE
```

Plot the time series for NEP at the monthly scale.  Be sure to add a horizontal 
line at 0 so that it is easy to see when NEP is positive or negative.

```{r}
#INSERT CODE
```

Plot the time series for NEP at the hourly scale.  Be sure to add a horizontal 
line at 0 so that it is easy to see when NEP is positive or negative.

```{r}
#INSERT CODE
```

Step 5: Were your hypotheses for NEP generally supported?

**Question 6:**  Hourly within a day

**Answer 6:**

**Question 7:** Monthly to month

**Answer 7:**

**Question 8:** Year to year

**Answer 8:**

## Part 3: Modeling the loss of carbon through respiration (RE)

Recall from the introduction that NEP is the net of two processes: 
Photosynthesis (GPP) and Respiration from plant and microbes (RE).  
To estimate GPP and RE we take advantage of the fact that we know GPP is zero at
night.

Step 1: Hypothesize how you expect air temperature to influence RE.

**Question 9:** Give your hypothesis for how RE varies with temperature.

**Answer 9:**

Step 2: Examine the relationship between NEP and temperature (`TA_F`) at night by 
using the NIGHT variable to select only the data that are identified as occurring
at NIGHT (see NIGHT column in the data) and plotting TA_F on the x-axis and NEP 
on the y-axis

```{r}
#INSERT CODE
```

Step 3: To model the relationship between air temperature and night time NEP (which 
is the same as night time RE) we will use a non-linear exponential model that 
looks like this:

```{r}
m <- nls(NEP ~ a * b^TA_F, night_only)
```

The object `m` is the output of the model.  It is the same type of object as the
liner regression models (`fit <- lm(y ~ x)`) we used in previous modules.  The 
difference is that we can use a non-linear model when we use the nls() function
to find the model. 

We can use the model object to make predictions and calculate residuals.  The 
`add_predictions` function takes the coefficients (or parameters) from the model
and creates predictions for each row.  It creates a new variable called `pred`. 
Similarly, `add_residuals()` calculates the difference between the observed data 
and the model (similar to Module 3).

```{r}
night_only <- night_only %>% 
  add_predictions(m) %>% 
  add_residuals(m)
```

Examine our model by adding the predicted nighttime NEP to the plot with the 
data as a colored line. This will use geom_points for plotting the data and 
geom_line for plotting the results from the model.  Be sure to look at the 
column name for the predictions from the model.

```{r}
#INSERT CODE
```

**Question 10:** Was your hypothesis for how RE varies with temperature 
supported by the data?  

**Answer 10:** 

Examine the residuals of the predicted NEP using a plot in the same way that we 
examined residuals in Module 3.

**Question 11:** Show the plot.  Are they normally distributed around 0?

**Answer 11:** 

```{r}
#INSERT CODE
```

Step 4: We can use this relationship between air temperature and NEP at night to 
calculate RE as a function of air temperature throughout the day. First, create a
function to calculate RE at all time periods in the full data set.  
Being able to make you own functions is a powerful way to develop code that is 
reusable.  For example, the following create a simple function that adds two 
number together. 

```{r}
sum_two_numbers <- function(a, b){
  c = a + b
  return(c)
}
```

Creating the function does not do anything other than defining the function.  
You need to call the function to use it

```{r}
sum_two_numbers(3, 1)
```

Now we can create a function that uses the output of the nls to calculate RE for
the full dataframe.  A positive value of RE is a loss of carbon from the 
ecosystem to the atmospheric through respiration.

```{r}
#df: is the data frame that gets automatically passed from the pipe
#m: is the model object (from lm or nls)
calculate_Re <- function(df, m){ 
  df <- df %>% add_predictions(m) %>% 
    rename("RE" = pred) %>% 
    mutate(RE = -RE) 
  return(df)
}
```

Note that `df` is the data frame that gets automatically passed from the pipe, 
and `m` is the model object from `lm` or `nls` function call.  If you are using 
the function without a pipe you need to use the command 
`calculate_Re(df = night_only, m = m)` but if you are using the function with a 
pipe then you can skip the df augument and use `calculate_Re(m = m)`

Now apply the function to all the flux data (do not rerun the nls function - you
need the nls function output from only the nighttime data because that is how RE
is calculated).  Remember that piping means that you do not need to include the 
`df` when using the function. 

```{r}
#INSERT CODE
```

Plot RE at the monthly time scale.  

```{r}
#INSERT CODE
```

**Question 12:** Do the patterns of RE at the monthly time-scale match your 
expectations from Question 3?

**Answer 12:**

## Part 4: Modeling the gain of carbon through respiration (GPP)

Now that we have an estimate to for RE during the date we can use the fact that 
NEP = GPP - RE (assuming, as we have done in our calculate_Re() function, that 
RE is a positive number if carbon is lost from the ecosystem) to estimate GPP 
for each time period.

```{r}
#INSERT CODE
```

Plot GPP at the hourly time scale.  

```{r}
#INSERT CODE
```

**Question 13:** Do the GPP patterns at the hourly time-scale match your 
expectations from Question 2?

**Answer 13:**

Finally, create a plot with the **monthly** values of NEP, GPP, and RE 
on the same plot (x-axis in month and y-axis is the flux value).  Do to this 
easily in ggplot I recommend using the `pivot_longer()` function to create a 
variable that is the type of flux (GPP, NEE, RE) and a column for the value

```{r}
#INSERT CODE
```

## Part 5: Analyze a new site

Pick a new site in a different eco-region and compare NEP values at the monthly 
and yearly scales.  Hint: you will need to use a join by month to combine the 
monthly summarized data from the two sites.

**Question 14:** What site did you chose?

**Answer 14:**

**Question 15:**

Use a plot to show how monthly NEP dynamics differ between the two sites 

**Answer 15:**
```{r}
#INSERT CODE
```

**Question 16:**

Make a plot of the annual NEP vs. year with both sites on the same figure.

**Answer 16:**
```{r}
#INSERT CODE
```

**Question 17:**
At the annual scale, which site removes more carbon from the atmosphere?

**Answer 17:**


## References

Components of the module are from:

Gough, C. M., Giffen, C., Woodward, T. W. (2018). Environmental Drivers of 
Ecosystem Carbon Fluxes from Minutes to Years. NEON Faculty Mentoring Network, 
QUBES Educational Resources. doi:10.25334/Q4VD7W
